Reproducible Codes

### 2020-10-20 Reproducible_codes.Rmd
adding revised files

---
title: "Supplementary Materials And Methods"
author: 
- Selcen Ari
- Alper Yilmaz 
date: "`r Sys.Date()`"
output:
  pdf_document: default
  html_document: default
  word_document: default
bibliography: NAR-sample.bib
header-includes: 
  \usepackage{float} \floatplacement{figure}{H} 
  \newcommand{\beginsupplement}{\setcounter{table}{0}  \renewcommand{\thetable}{S\arabic{table}} \setcounter{figure}{0}     \renewcommand{\thefigure}{S\arabic{figure}}}
  \usepackage{subfig}
---
\tableofcontents

@@

# 1. Functions defined for ceRNA models and workflow of method

We defined the functions that can be used with R programming. Briefly, these functions process a given miRNA:gene dataset and convert to graph object. All values that are significant in miRNA:target interactions are stored in edge variables and processed with formulations that are given in previous section. The functions and steps of approach are explained as following (Figure S1 in Supplementary Figures) :

**Conversion of dataset**: `priming_graph()` function processes the given dataset that includes competing elements in first variable and repressive element in second variable. If the affinity and/or degradation factors are specified in the function, factors are taken into account, are processed with defaults in vice versa. The formulations that are given in equations (1-4) are performed in this function. This step gives the graph object which contains efficiency values of miRNA:competing target pairs in steady-state in terms of amount. It is assumed that the initial target amounts in the dataset is observed after the repressive activity of miRNAs in steady-state.

**Transition of variables in graph**: In the previous step, the calculations are performed in the edge variables of the graph object. However, the graph object allows to use node variables, while the node features are handled to the graph. In this direction, update_nodes function carries the amount values to node variables This step must be applied with “once” option because it is primary process.

**Trigger change in graph**: The dataset are assumed as steady-state in previous step and the efficiency coefficients are calculated according to this acceptance. In the network that is found in steady-state conditions, the change is applied to the graph object for perturbation of steady-state. To provide the perturbation in the network the work-flow offer two methods: *update_variables* and *update_how*. The first, a new dataset that is contained competing and repressive element names and current values of these can be processed with update_variables. The second option, the amount of the given node name in update_how function can be changed according to “how” argument.

**Simulation of competing behavior of targets**: After the change in the steady-state conditions, the network elements try to gain steady-state again. This process progresses as repeating of regulations after the spreading the changes in the network. In this step, simulation of regulations according to given cycle count in simulate function is applied. After each simulation cycle, the miRNA repression values are re-calculated and the current values of competing elements are found and saved. The process is performed in the edge data and at the same time outputs of the calculations are carried from edge to node data.

The node elements in the dataset are handled as two type; repressive (miRNAs) and competing (targets). It is assumed in approach that while targets are degrading or inhibiting by miRNAs continuously, miRNAs reversibly used. If the trigger of the network is a miRNA, it maintains the current value of amount that provides by user. On the contrary, it tries to help this process to provide steady-state through the regulations on its amount, if a competing element is used as a trigger. The functions that are used in the this study are developed with R and are available in [Bioconductor](https://www.bioconductor.org/packages/release/bioc/html/ceRNAnetsim.html).

@@

+ load *minsamp* data

```{r}
data("minsamp")
minsamp
```

@@

# 2. *minsamp* dataset analysis in lack of interaction factors.

@@

# 3. *minsamp* dataset analysis with interaction factors.

We have made the same analysis in presence of interaction factors (Sequentially shown at Figure S5-7 in Supplementary Figures).

@@

# 4. Common target perturbation in *minsamp* dataset.

Genes targeted by multiple miRNAs (referred to as "common target") are of special interest since they are subject to cooperative effect. Also, they perturb more than one neighborhood. In our small dataset, minsamp, Gene4 is regulated by two miRNAs. Let's simulate perturbation effects triggered by Gene4 (Shown at Figure S8 in Supplementary Figures) .

@@

# 5. Determination of perturbation efficiencies of elements in system.

priming_graph(minsamp, competing_count = Competing_expression, miRNA_count = miRNA_expression, aff_factor = c(energy, seed_type), deg_factor = region)-> sample_graph

find_node_perturbation(sample_graph, how = 2, cycle = 3, limit = 0.1)%>%
  as_tibble()%>%
  select(name, perturbation_efficiency, perturbed_count)

# 6. Additional data manipulation steps

## 6.1 Arrangement of CLASH dataset

CLASH dataset was retrieved from [PubMed](https://www.ncbi.nlm.nih.gov/pmc/articles/PMC3650559/) [@helwak_mapping_2013].

```{r, eval=FALSE}

clashelwak <- read.table("mmc1.txt", comment.char = "#", header = TRUE, skip = 1, stringsAsFactors = FALSE)

#hg19
```

### Query of Human Genome 19.

Human genome 19 information was handled through biomaRt package.

```{r,  eval=FALSE}
#HG19
listEnsemblArchives()
listMarts(host = 'http://grch37.ensembl.org')
ensemblgrch37 = useMart(host='http://grch37.ensembl.org', 
                     biomart='ENSEMBL_MART_ENSEMBL',
                    dataset='hsapiens_gene_ensembl')
hg19 <- getBM(attributes = c("ensembl_transcript_id","ensembl_gene_id","chromosome_name", "start_position","end_position", "hgnc_symbol", "entrezgene_id", "strand" ), mart = ensemblgrch37)

### Adding miRNA and gene information

```{r,  eval=FALSE}

clashelwak <- clashelwak%>%
 separate(microRNA_name, c("Barcode", "Database", "mirna_name", "type"), sep = "_")%>%
  separate(mRNA_name, c("Ensembl_Gene_Id", "Ensembl_Transcript_Id", "Hugo_Symbol", "mRNA_Type"), sep = "_")

MiRNA releases are obtained from [miRBase](http://www.mirbase.org/ftp.shtml). In this step, release 21 (in Human genome 38) was downloaded.

```{r, eval=FALSE}
read.table("mirbasehg38.txt", comment.char = "#")%>%
  filter(V3 != "miRNA_primary_transcript")%>%
  separate(V9, c("ID", "Alias", "Name", "Precusor"), sep = ";")%>%
  mutate(ID = substr(ID, 4, length(ID)), Alias = substr(Alias, 7, length(Alias)), Name = substr(Name, 6, length(Name)), Precusor = substr(Precusor, 14, length(Precusor)))%>%
  dplyr::select(chr= V1, start = V4, end = V5, strand = V7, ID, Alias, Name, Precusor)->mirbasehg38

CLASH dataset is published in miRBase release 15 and Human Genome 19 version.

```{r, eval=FALSE}
read_tsv("mirna_mature.txt", col_names = FALSE)%>%
  filter(startsWith(X2, "hsa"))%>%
  dplyr::select(mirna_ID = X2, mirbase_ID = X3)%>%
  inner_join(mirbasehg38%>% dplyr::select(ID, Name), by= c("mirbase_ID"= "ID"))%>%
  dplyr::select(mirbase_ID, Name)%>%
  distinct()%>%
  inner_join(clashelwak, by = c("mirbase_ID"= "Barcode"))%>%
  dplyr::select(Name, miRNA_seq, Ensembl_Gene_Id, Ensembl_Transcript_Id, Hugo_Symbol, mRNA_seq_extended, chimeras_decompressed, seed_type, seed_basepairs,folding_class, seq_ID, folding_energy, X5.UTR, CDS, X3.UTR)%>%
  inner_join(hg19, by = c("Ensembl_Gene_Id"= "ensembl_gene_id", "Ensembl_Transcript_Id"="ensembl_transcript_id", "Hugo_Symbol"="hgnc_symbol"))%>%
  mutate(region1 = ifelse(X5.UTR == "1", "5UTR", " "), region2= ifelse(X3.UTR == "1", "3UTR", " "),region3= ifelse(CDS =="1", "CDS", " "))%>%
  unite(region, c(region1, region2, region3), sep = "||")%>%
  dplyr::select(chromosome_name, start_position, end_position,strand, Hugo_Symbol, Ensembl_Gene_Id, Ensembl_Transcript_Id, mRNA_seq_extended, Name, miRNA_seq, seq_ID, seed_type,seed_basepairs, folding_class, folding_energy, region)%>%
  as_tibble() -> clashelwakfinal

### Converting CLASH data to human genome 38 build.

There are different liftover methods for conversion among Human Genome builds. We preffered to use [UCSC liftover tool](https://genome.ucsc.edu/cgi-bin/hgLiftOver)

```{r,  eval=FALSE}

#Obtaining chromosomal locations from miRNA:target interaction dataset.

lift19 <- clashelwakfinal%>%
  dplyr::select(1,2,3)%>%
  unite(start_end, c("start_position", "end_position"), sep = "-")%>%
  mutate(Chromosome = paste0("chr",chromosome_name, ""))%>%
  unite(chromosome_name, c("Chromosome","start_end"), sep = ":")

write_tsv(lift19, "lift19.txt")

#After we searched this file in UCSC browser, the output loaded (lift19_del deleted regions on the HG38 genome build; hg38clashcomp new locations of the genes on HG38)

lift19_del <- read_tsv("deleted_lift19.txt")
colnames(lift19_del)[1] <- "chromosome_loc"

lift19_del <- lift19_del%>%
  dplyr::filter(startsWith(chromosome_loc, "chr"))%>%
  separate(chromosome_loc, c("Chr", "End"), "-", remove = TRUE)%>%
  separate(Chr, c("Chr", "Start"), ":", remove = TRUE)

lift19_del$Start <- as.numeric(lift19_del$Start)
lift19_del$End <- as.numeric(lift19_del$End)

#removing deleted location from CLASH dataset

clashelwakfinal <-clashelwakfinal%>%
  mutate(Chromosome = paste0("chr",chromosome_name, ""))%>%
  dplyr::anti_join(lift19_del, by = c("Chromosome"="Chr", "start_position"="Start", "end_position"="End"))

hg38clash <- read.delim("hg38clashcomp.txt", header = FALSE, stringsAsFactors = FALSE)

# adding new location information:

clashelwakfinal <- clashelwakfinal%>% 
  bind_cols(hg38clash)

colnames(clashelwakfinal)[18] <- "HG38build_loc"

clashelwakfinal <- clashelwakfinal%>%
  dplyr::mutate(Genom_build= rep("hg19"))

# Arrangement in dataset

clashelwakfinal%>% dplyr::select(cluster=seq_ID, chromosome = Chromosome, start_position, end_position, strand, hgnc_symbol = Hugo_Symbol, Ensembl_Gene_Id, Ensembl_Transcript_Id, target_seq = mRNA_seq_extended, miRNA= Name, miR_seq =miRNA_seq, seed_type, seed_type2= seed_basepairs, seed_type3= folding_class, Energy = folding_energy, HG38build_loc, Genom_build, region) -> clashelwakfinal

clashelwakfinal$strand <- as.character(clashelwakfinal$strand)

str(clashelwakfinal)
```

### Interpreting the CLASH seed structures in dataset

```{r,  eval=FALSE}
clashelwakfinal%>%
  mutate(seed_type= ifelse (seed_type == "noncanonical_seed"& seed_type2 >4 &seed_type3=="I", paste0(seed_type2, "-mer"), seed_type),
         seed_type= ifelse (seed_type == "noncanonical_seed"& seed_type2 >4 &seed_type3=="II", paste0(seed_type2, "-mer_noncanonical"), seed_type), 
         seed_type= ifelse (seed_type == "noncanonical_seed"& seed_type2 >4 &seed_type3=="III", paste0(seed_type2, "-mer_noncanonical"), seed_type),
         seed_type= ifelse (seed_type == "noncanonical_seed"& seed_type2 >4 &seed_type3=="IV", paste0(seed_type2, "-mer_noncanonical"), seed_type),
         seed_type= ifelse (startsWith(seed_type, "no") ,"none", seed_type))%>%
  dplyr::select(-seed_type2, -seed_type3)->clashelwakfinal

## 6.2 Arrangement of CLEAR-CLiP Dataset [@moore_mirnatarget_2015]

CLASH dataset was retrieved from [Nature article](nature.com/articles/ncomms9864)

```{r,  eval=FALSE}
clearclip <- read_xlsx("CLEAR-CLIP.xlsx")

#Clearclip hg18
```

### Query of Human Genome 18

```{r,  eval=FALSE}
#HG18
listEnsemblArchives()
listMarts(host = 'may2009.archive.ensembl.org')
ensembl54 = useMart(host='may2009.archive.ensembl.org', 
                     biomart='ENSEMBL_MART_ENSEMBL',
                    dataset='hsapiens_gene_ensembl')

hg18 <- getBM(attributes = c("ensembl_transcript_id","ensembl_gene_id","chromosome_name", "start_position","end_position", "hgnc_symbol", "entrezgene", "strand" ), mart = ensembl54)
```

### Adding Genome Information to dataset

```{r,  eval=FALSE}
clearclipfinal <- hg18%>%
  inner_join(clearclip, by= c("entrezgene"= "gene.id", "hgnc_symbol"= "gene.symbol"))%>%
  distinct()
```

### Converting human genome build

```{r,   eval=FALSE}

#Obtaining chromosomal locations from miRNA:target interaction dataset.

lift18 <- clearclipfinal%>%
  unite(start_end, c("start_position", "end_position"), sep = "-")%>%
  unite(location, c("chr","start_end"), sep = ":")%>%
  dplyr::select(location)

write_tsv(lift18, "lift18.txt")

#After we searched this file in UCSC browser, the output loaded (deleted_lift18 deleted regions on the HG38 genome build; hg38clearclip new locations of the genes on HG38)

deleted_lift18 <- read_tsv("deleted_lift18.txt")

colnames(deleted_lift18)[1] <- "Chromosome_loc"

deleted_lift18 <- deleted_lift18%>%
  dplyr::filter(startsWith(Chromosome_loc, "chr"))%>%
  separate(Chromosome_loc, c("Chr", "End"), "-", remove = TRUE)%>%
  separate(Chr, c("Chr", "Start"), ":", remove = TRUE)

deleted_lift18$Start <- as.numeric(deleted_lift18$Start)
deleted_lift18$End <- as.numeric(deleted_lift18$End)

#removing deleted location from CLEAR-CLiP dataset

clearclipfinal <- clearclipfinal%>%
  dplyr::anti_join(deleted_lift18, by = c("chr"="Chr", "start_position"="Start", "end_position"="End"))

hg38clearclip<- read.delim("hg38clearclip.txt", header = FALSE, stringsAsFactors = FALSE)

clearclipfinal <- clearclipfinal%>% 
  bind_cols(hg38clearclip)

colnames(clearclipfinal)[28] <- "HG38build_loc"

# adding new location information:

clearclipfinal <- clearclipfinal%>%
  dplyr::mutate(Genom_build= rep("hg18"))

# Arrangement in dataset

clearclipfinal%>%
  dplyr::select(cluster=cluster.ID, chromosome = chr, start_position, end_position, strand = strand.y, hgnc_symbol, Ensembl_Gene_Id =ensembl_gene_id, Ensembl_Transcript_Id =ensembl_transcript_id, target_seq = target.map, miRNA, miR_seq = miR.map, seed_type = 'seed match', Energy = MFE, HG38build_loc, Genom_build, region) -> clearclipfinal

### Seed type manipulation in CLEAR-CLiP dataset

In CLEAR-CLiP dataset, seed types were shown in detail. We adjusted as canonical and non-canonical.

```{r,  eval=FALSE}
data_frame(seed_type= c("5mer_1","5mer_2", "5mer_3", "6mer", "6mer.indel", "6mer.mm", "6mer_off.mm", "6merA1", "6merA1.indel", "6merA1.mm", "7merA1", "7merA1.indel", "7merA1.mm", "7merm8", "7merm8.indel", "7merm8,mm", "8mer", "8mer.indel", "8mer.mm", "NA" ),
           seed_type_com= c("5-mer", "5-mer_noncanonical", "5-mer_noncanonical", "6-mer", "6-mer_noncanonical", "6-mer_noncanonical", "6-mer_noncanonical", "6-merA1", "6-merA1_noncanonical","6-merA1_noncanonical", "7-merA1", "7-merA1_noncanonical","7-merA1_noncanonical", "7-mer-8m", "7-mer-8m_noncanonical", "7-mer-8m_noncanonical", "8-mer", "8-mer_noncanonical", "8-mer_noncanonical", "none" )) ->clipdata_seed

clearclipfinal%>%
  inner_join(clipdata_seed, by = "seed_type")%>%
  dplyr::select(1:11, seed_type = seed_type_com, Energy, HG38build_loc, Genom_build, region)-> clearclipfinal

## 6.3 Integration of two experimental dataset

```{r,  eval=FALSE}

bind_rows(clashelwakfinal, clearclipfinal)%>%distinct() -> experimentalmirnagene

### Adding Coefficients of Interaction factors

Energy values in miRNA:target pairs are represented by high-throughput studies [@helwak_mapping_2013; @moore_mirnatarget_2015] which are utilized in this study. On the other hand, we have specified the other interaction factors, seed type and location of binding region on the target, as numeric values based on the previous studies.[@grimson_microrna_2007] have compared the seed types' effect on target repression with few miRNA had canonical seed pairing in their study. Additionally, [@bartel_micrornas:_2009] and [@betel2010comprehensive] have studied on functional and non-functional seed interactions. Based on results of these studies we have arranged seed types of miRNA:target interactions as numeric values. We also have redefined location of binding region on the target as numeric values, based on studies of [@hausser_analysis_2013] and [@helwak_mapping_2013]. With this process, we have handled this integrated dataset in context of competitor behaviors and functionality of interactions.

In this step we added numeric intraction values at followings

Fistly, we organized these values due to the fact that the regions were defined differently in two datasets. After that, region effect was added as numeric values (shown in Table S3).

```{r,  eval=FALSE}
experimentalmirnagene <- experimentalmirnagene%>%
  mutate(region2  = str_replace_all(region, "NA", ""), region3 = str_replace_all(region2, "\\|", ""), region = str_replace_all(region3, c("3'UTR"="3UTR", "5'UTR"="5UTR")))%>%
  dplyr::select(-region2, -region3)%>%
  mutate(region_effect = as.double(ifelse(region %in% c("3UTRCDS", "CDS3UTR", "5UTR3UTR", "CDS5UTR3UTR", "CDS3UTRintron"), "0.93", ifelse(region %in% c("CDS", "CDSintron"), "0.42", ifelse(region %in% c("3UTR","3UTRintron"), "0.84", ifelse(region %in% c("5UTR", "5UTRintron"), "0.01", ifelse(region %in% c("5UTRCDS", "CDS5UTR"), "0.42", ifelse(region %in% c("intron",""), "0.01", ifelse(region %in% c("exon_unclassified", ""), "0.2", NA)))))))))

Secondly, we organized seed type interactions in *Seed type manipulation* section for CLEAR-CLiP dataset to show as found in CLASH dataset. Same type formatted values added dataset as numeric values (shown in Table S2).

```{r,  eval=FALSE}

seed_type_effect <- data_frame( seed_type = c("5-mer", "5-mer_noncanonical", "6-mer", "6-mer_noncanonical", "6-merA1", "6-merA1_noncanonical", "7-mer", "7-mer_noncanonical", "7-merA1", "7-merA1_noncanonical", "7-mer-8m", "7-mer-8m_noncanonical", "8-mer", "8-mer_noncanonical", "9-mer", "9-mer_noncanonical", "none"),
                                seed_type_effect= c(0.05, 0.04, 0.07, 0.05, 0.07, 0.05, 0.23, 0.19, 0.19, 0.16, 0.25, 0.21, 0.43, 0.35, 0.43,0.35, 0.01)
                                )

experimentalmirnagene%>%
  inner_join(seed_type_effect, by= "seed_type") -> experimentalmirnagene
```

```{r}
readRDS("data/experimentalmirnagene.RDS")->experimentalmirnagene
experimentalmirnagene
```

The context of dataset is shown in Table S5 in Supplementary Tables.

# REFERENCES

>>>>

---
title: "Reproducible Codes"
author: 
- Selcen Ari
- Alper Yilmaz 
date: "`r Sys.Date()`"
output:
  pdf_document: default
  html_document: default
  word_document: default
bibliography: NAR-sample.bib
header-includes: 
  \usepackage{float} \floatplacement{figure}{H} 
  \newcommand{\beginsupplement}{\setcounter{table}{0}  \renewcommand{\thetable}{S\arabic{table}} \setcounter{figure}{0}     \renewcommand{\thefigure}{S\arabic{figure}}}
  \usepackage{subfig}
---
\tableofcontents

@@

@@

```{r, include=FALSE}
data("minsamp") # refers Sample and Sample+ data
```

@@

# 2. Sample dataset analysis in lack of interaction factors.

@@

# 3. Sample+ dataset analysis with interaction factors.

We have made the same analysis in presence of interaction factors (Sequentially shown at Figure S6-8 in Supplementary Figures).

@@

# 4. Common target perturbation in Sample+ dataset.

Genes targeted by multiple miRNAs (referred to as "common target") are of special interest since they are subject to cooperative effect. Also, they perturb more than one neighborhood. In our small dataset, minsamp, Gene4 is regulated by two miRNAs. Let's simulate perturbation effects triggered by Gene4 (Shown at Figure S9 in Supplementary Figures) .

@@
